Revisiting the Austral Spring Extratropical Southern Hemisphere zonally assymetric circulation using complex Empirical Orthogonal Functions

### 2021-09-07 analysis/paper/shceof.Rmd
Cambia datos y hace muchos cambios al texto.

```{r setup, echo = FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  cache = TRUE,
  comment = "#>",
  cache.extra = 41,
  cache.path = "../cache/paper/",
  fig.path = "../figures/",
  dpi = 300,
  dev = c("pdf", "png")
)

@@

era_more %>%
  copy() %>% 
  .[, ref := year(time) > 1979] %>% 
  # .[, hgt := scale(hgt, center = FALSE, scale = sd(hgt[ref])), by = .(part, cEOF)] %>%
  ggplot(aes(time, hgt*1e3)) +
  # Zero line + rangeframe
  geom_hline(data = ~.x[, mean(hgt)*1e3, by  = .(cEOF, part)], 
             aes(yintercept = V1), size = 0.2, color = "gray30") +
  ggthemes::geom_rangeframe(sides = "r", size = 0.2, color = "gray30") +
  geom_line(aes(color = part)) +
  geom_point(aes(color = part), size = 0.25) +
  # geom_line(data = era_less) +
  scale_alpha_manual(values = c(1, 0.2)) +
  geom_smooth(method = "lm", aes(color = stage(part, after_scale = scales::muted(color))),
              show.legend = FALSE, size = 0.8) +
  scale_x_datetime(NULL, date_labels = "%Y", 
                   breaks = seq(as.POSIXct("1950-01-01"), 
                                as.POSIXct("2020-01-01"), by = "10 years"),
                   minor_breaks = seq(as.POSIXct("1950-01-01"), 
                                      as.POSIXct("2020-01-01"), by = "2 years")) +
  scale_y_continuous(NULL, minor_breaks = NULL, sec.axis = dup_axis()) +
  scale_color_brewer(NULL, palette = "Dark2", 
                     guide = guide_legend(override.aes = list(fill = NA))) +
  facet_grid(cEOF ~ part) +
  theme(panel.spacing = grid::unit(1.5, "lines"),
        strip.placement = "outside", 
        panel.grid.minor.x = element_line(linetype = 2))  +
  tag_facets("cr") +
  panel_background 
```

@@

```{r regr_psi}
regr_psi <- ceof[, eof[[1]]$left, by = .(season)] %>%
  sep_ReIm(hgt, "wide") %>%
  .[, hgt := NULL] %>%
  .[, `:=`(Real = Real/sd(Real), Imaginary = Imaginary/sd(Imaginary)), 
    by = .(season, cEOF)] %>%
  .[psi, on = "time", allow.cartesian = TRUE] %>%
  .[lev == 200] %>% 
  .[season(time) == "SON"] %>%
  .[, FitLm(psi.z, Real, Imaginary, se = TRUE),by = .(lon, lat, lev, cEOF, season)] %>%
  rm_intercept() %>%
  .[, psi.z := Anomaly(estimate), by = .(cEOF, season, lev, lat)] %>%
  .[, c("fx", "fy") := shceof::WaveFlux(.SD, p = 200), by = .(lev, cEOF, season, term)] %>%
  .[, term := factor_ReIm(term)]
```

@@

```{r regr_sst}
regr_sst <- ceof[, eof[[1]]$left, by = .(season)] %>%
  sep_ReIm(hgt, "wide") %>%
  .[, hgt := NULL] %>%
  .[, `:=`(Real = Real/sd(Real), Imaginary = Imaginary/sd(Imaginary)), 
    by = .(season, cEOF)] %>%
  .[sst, on = "time", allow.cartesian = TRUE] %>%
  .[season(time) == "SON"] %>%
  .[, FitLm(t, Imaginary, Real, se = TRUE), by = .(lon, lat, cEOF, season)] %>%
  rm_intercept() %>%
  .[, term := factor_ReIm(term)]

```{r enso_cor}
enso_cor <-  ceof[, eof[[1]]$left, by = .(season)] %>%
  sep_ReIm() %>%
  .[, hgt := scale(hgt), by = .(cEOF, part)] %>%
  enso[., on = "time"] %>%
  na.omit() %>%
  .[, correlate(hgt, oni), by = .(cEOF, part)]

```{r plot_sst_psi}
plot_sst_psi <- function(regr_sst, regr_psi, which_cEOF, which_lev) {
  sst_gdata <- regr_sst[cEOF == which_cEOF] %>%
    .[lat %between% c(-90, 10)] %>%
    .[, p_val := Pvaluate(estimate, std.error, df, "fdr"), 
      by = .(cEOF, term)] %>%
    .[, var := "SST"]

  rsst <- max(range(abs(regr_sst[!is.na(cEOF), ]$estimate), na.rm = TRUE))
  breaks_sst <- AnchorBreaks(0, bins = 20, exclude = 0)(c(-rsst, rsst))

  stream_gdata <- regr_psi %>%
    .[cEOF == which_cEOF] %>%
    .[lev == which_lev] %>%
    # .[abs(estimate) > 1, estimate := NA] %>%
    .[lat %between% c(-90, 10)] %>%
    .[, p_val := Pvaluate(estimate, std.error, df, "fdr"), by = .(cEOF, term)] %>%
    .[, var := "Streamfunction"]

  rstream <- max(range(abs(regr_psi[lev == which_lev]$estimate), 
                       na.rm = TRUE))*1e-7
  breaks_stream <- AnchorBreaks(0, bins = 20, exclude = 0)(c(-rstream, rstream))

  # browser()
  stream_gdata %>% 
    periodic(lon = c(0, 360)) %>%
    ggplot(aes(lon, lat)) +
    geom_contour_fill(aes(z = estimate*1e-7, fill = ..level..),
                      breaks = breaks_stream) +
    geom_contour_pval(aes(z = p_val)) +
    geom_streamline(aes(dx = fx, dy = fy), min.L = 3, L = 10,
                    size = 0.1, res = 2, nx = 22, ny = 12, 
                    arrow.angle = 14, arrow.length = 0.2) +
    geom_qmap(keep = .015) +
    scale_fill_divergent_discretised(latex2exp::TeX(r'(Streamfunction ($m^2s^{-1}\times 1e^{-7}$))'), 
                                     labels = one_less,
                                     low = scales::muted("red"),
                                     high = scales::muted("blue"),
                                     guide = guide_colorsteps_bottom(15, show.limits = FALSE,
                                                                     order = 99)) +
    ggnewscale::new_scale_fill() +

    geom_contour_fill(data = sst_gdata,
                      aes(z = estimate, fill = ..level..), 
                      breaks = breaks_sst) +
    geom_contour_pval(data = sst_gdata, aes(z = p_val)) +
    scale_fill_divergent_discretised("SST (K)", 
                                     labels = one_less,
                                     guide = guide_colorsteps_bottom(width = 15,
                                                                     show.limits = FALSE, ticks = TRUE,
                                                                     order = 1)) +
    scale_x_continuous(NULL, expand = c(0, 0), 
                       breaks = seq(-180, 180-60, by = 60),
                       labels = LonLabel) +

    scale_y_continuous(NULL, breaks  = seq(-90, 0, by = 30), expand = c(0, 0)) +
    scale_mag(guide = "none") +
    facet_grid(term ~ var) +
    # coord_quickmap(ylim = c(NA, 10)) +
    coord_sf(ylim = c(-89, 0), expand = FALSE,
             crs = "+proj=longlat +pm=180 +over",
             default_crs = "+proj=longlat +lon_wrap=180 +over") + 
    theme(legend.box = "horizontal",
          legend.text = element_text(size = rel(0.7)), 
          legend.box.spacing = grid::unit(0, "lines") ) +
    tag_facets("cr", position = "bl") 
}

(ref:sst-psi-2-cap) Regression maps of cEOF2 with SST (K, column a) and streamfunction zonal anomalies ($m^2/s\times10^-7$, shaded) with their corresponding activity wave flux (vectors) (column b) for the 1979 -- 2019 period. Areas marked with dots have p-values smaller than 0.01 adjusted for FDR.

```{r sst-psi-2,  fig.cap = "(ref:sst-psi-2-cap)", fig.width = 7, fig.height=4, dependson="plot_sst_psi", fig.env="figure*"}
plot_sst_psi(regr_sst = regr_sst, regr_psi = regr_psi,
             which_cEOF = "cEOF2", which_lev = 200) 
```

The connections between cEOFs and Tropical sources are also assessed.
Figure \@ref(fig:sst-psi-2) shows the regression maps of Sea Surface Temperatures (SST) and Streamfunction anomalies at 200 hPa respectively upon cEOF2. The Imaginary cEOF2 is associated with strong positive SST anomalies on the Central Pacific and negative anomalies over an area across the North of Australia and New Zealand, the South Pacific Convergence Zone (SPCZ) (Figure \@ref(fig:sst-psi-2).a2). This pattern is almost canonically positive ENSO [@bamston1997] and indeed, the correlation between the Imaginary cEOF2 and the Oceanic Niño Index [@bamston1997] is significant and very high `r enso_cor[cEOF == "cEOF2" & part == "Imaginary"]$text`.
Besides, the Imaginary cEOF2 is associated with strong wave-like streamfunction anomalies emanating from the tropics (Figure \@ref(fig:sst-psi-2).b2). This is consistent with the effect of ENSO on the extratropics: SST anomalies initiate anomalous tropical convection that excites Rossby waves that propagate meridionally towards higher latitudes [@mo2000].

Since the Real cEOF2 represents just a different phase of the same wave train, it would be expected to show a similar forcing pattern to the Imaginary cEOF2 with a slight translation of its location.
However, Figure \@ref(fig:sst-psi-2).a1 and b1 show that the Real cEOF2 is not associated either with any significant SST nor streamfunction anomalies in the tropics.
The correlation between the Real cEOF2 and ENSO is also not significant (`r enso_cor[cEOF == "cEOF2" & part == "Real"]$text`).
This lack of tropical signal suggest that the Real cEOF2 and the Imaginary cEOF2 have very different sources of variability.

(ref:enso-phase-cap) ONI plotted against cEOF2 phase for th 1979 -- 2019 period.
Years with magnitude of cEOF2 greater or smaller than the 50th percentile are shown as orange circles and green diamomnds respectively. 
Black line is the fit ONI ~ sin(phase) computed by weighted OLS using the magnitude of the cEOF2 as weights.

```{r enso-phase, fig.cap = "(ref:enso-phase-cap)", fig.width = width_column, fig.height=3}
gdata <- ceof[, eof[[1]]$left, by = .(season)] %>%
  .[cEOF == "cEOF2"] %>% 
  # .[, hgt := Anomaly(hgt)] %>% 
  .[, arg := Arg(hgt)] %>%
  .[, mag := Mod(hgt)] %>%
  .[enso, on = "time"] %>%
  na.omit() %>%
  .[, q := ifelse(mag >= quantile(mag, 0.5), "|cEOF2| > 50%", "|cEOF2| < 50%")]

@@

  geom_smooth(data = ~qwrap(.x, arg = c(-pi, pi) ~ c(-2*pi, 2*pi)),
              method = "lm", aes(weight = mag),
              formula = y ~ I(sin(x)) -1,
              se = FALSE, color = "black", size = 0.5) +
  geom_point(aes(color = q,
                 shape  = q),
             size = 3) +

@@

Furthermore, Figure \@ref(fig:enso-phase) suggest that highest amplitude cEOF2 years tend to coincide with strong ENSO years.
The correlation between the absolute magnitude of the ONI and the magnitude of the cEOF2 is `r cor_enso_mag$text`.
This relationship, however, appears to be driven only by the three years with strongest ENSO events in the period (`r knitr::combine_words(year(gdata[order(-oni)][1:3]$time))`) which also coincide with the three years with strongest cEOF2 magnitude.
If those years are removed, the correlation becomes non-significant (`r cor_enso_mag_outlier$text`).
Furthermore, even when using all the datapoints, the Spearman correlation --which is robust to outliers-- is also non-significant (`r signif(cor_enso_spearman$estimate, 2)`, p-value`r report_p(cor_enso_spearman$p.value)`).
Therefore, the relationship between the magnitude of the cEOF2 train wave and ONI remains uncertain.

```{r}
# gdata %>% 
#   ggplot(aes(oni, mag)) +
#   geom_point(aes(color = mag >= quantile(mag, 0.5)))  +
#    scale_color_brewer(NULL, palette = "Dark2",
#                      labels = c("TRUE"  = "|cEOF2| > 50%",
#                                 "FALSE" = "|cEOF2| < 50%")) +
#   panel_background
```

It could be concluded that the wave train represented by cEOF2 can be both part of the internal variability of the extratropical atmosphere or forced by tropical SSTs.
In the former case, the wave train has little phase preference.
However, when cEOF2 is excited by tropical SST variability, it tends to remain locked to the imaginary phase.
This explains the relative overabundance of years with cEOF2 near positive and negative imaginary phase in Figure \@ref(fig:phase-histogram).

```{r sst-psi-1, fig.cap = "(ref:sst-psi-1-cap)", fig.width = 7, fig.height=4, dependson="plot_sst_psi", fig.env = "figure*"}
plot_sst_psi(regr_sst = regr_sst, regr_psi = regr_psi,
             which_cEOF = "cEOF1", which_lev = 200) 
```

Figure \@ref(fig:sst-psi-1) shows SST and streamfunction regression maps for cEOF1.
Unlike the cEOF2 case, there is no significant pattern of SST anomalies associated with either the Real or Imaginary cEOF1.
Consistently, streamfunction anomalies do not show any tropical influence.
Instead, the real and Imaginary cEOF1 are associated with zonally wave activity fluxes in the extra-tropics around 60ºS, except for an equatorward flow from the coast of Antarctica around 150ºE in the real component.
This suggests that it's variability is driven primary by the internal variability of the extra-tropics.

## Precipitation {#precipitation}

@@

```{r}
rotate_reim <- function(real, imaginary, angle = 0) {
  z <- complex(real = real, imaginary = imaginary)
  z <- rotate(z, angle)
  list(real = Re(z),
       imaginary = Im(z))
}

sst_for_regression <- ceof[, eof[[1]]$left, by = .(season)] %>%
  .[cEOF == "cEOF2"] %>% 
  copy() %>% 
  sep_ReIm(hgt, "wide") %>%
  .[, hgt := NULL] %>%
  .[sst, on = "time", allow.cartesian = TRUE] %>%
  .[season(time) == "SON"] %>% 
  na.omit()

angles_regr <- -c(0, 45, 90, 45+90)*pi/180

sst_regr <- lapply(angles_regr, function(a) {
  sst_for_regression %>% 
    copy() %>% 
    .[, c("Real", "Imaginary") := rotate_reim(Real, Imaginary, a)] %>% 
    .[, `:=`(Real = Real/sd(Real), Imaginary = Imaginary/sd(Imaginary)), 
      by = .(season, cEOF)] %>%
    .[, FitLm(t, Imaginary, Real, se = TRUE), by = .(lon, lat, cEOF, season)] %>%
    rm_intercept() %>%
    .[term == "Real"] %>% 
    .[, term := factor_ReIm(term)] %>% 
    .[, angle := a]
}) %>% 
  rbindlist()

psi_for_regression <-  ceof[, eof[[1]]$left, by = .(season)] %>%
  .[cEOF == "cEOF2"] %>% 
  copy() %>% 
  sep_ReIm(hgt, "wide") %>%
  .[, hgt := NULL] %>%
  .[psi, on = "time", allow.cartesian = TRUE] %>%
  .[season(time) == "SON"] %>%
  .[lev == 200]

>>>>

```{r setup, echo = FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  cache = TRUE,
  comment = "#>",
  cache.extra = 43,
  cache.path = "../cache/paper/",
  fig.path = "../figures/",
  dpi = 300,
  dev = c("pdf", "png")
)

@@

era_more %>%
  copy() %>% 
  .[, ref := year(time) > 1979] %>% 
  # .[, hgt := scale(hgt, center = FALSE, scale = sd(hgt[ref])), by = .(part, cEOF)] %>%
  ggplot(aes(time, hgt*1e3)) +
  # Zero line + rangeframe
  geom_hline(data = ~.x[, mean(hgt)*1e3, by  = .(cEOF, part)], 
             aes(yintercept = V1), size = 0.2, color = "gray30") +
  ggthemes::geom_rangeframe(sides = "r", size = 0.2, color = "gray30") +
  geom_line(aes(color = part)) +
  geom_point(aes(color = part), size = 0.25) +
  # geom_line(data = era_less) +
  scale_alpha_manual(values = c(1, 0.2)) +
  geom_smooth(method = "lm", aes(color = stage(part, after_scale = scales::muted(color))),
              show.legend = FALSE, size = 0.8) +
  scale_x_datetime(NULL, date_labels = "%Y", 
                   breaks = seq(as.POSIXct("1950-01-01"), 
                                as.POSIXct("2020-01-01"), by = "10 years"),
                   minor_breaks = seq(as.POSIXct("1950-01-01"), 
                                      as.POSIXct("2020-01-01"), by = "2 years")) +
  scale_y_continuous(NULL, minor_breaks = NULL, sec.axis = dup_axis()) +
  scale_color_brewer(NULL, palette = "Dark2", 
                     guide = guide_legend(override.aes = list(fill = NA))) +
  facet_grid(cEOF ~ part) +
  theme(panel.spacing = grid::unit(1.5, "lines"),
        strip.placement = "outside", 
        panel.grid.minor.x = element_line(linetype = 2))  +
  tag_facets("rc") +
  panel_background 
```

@@

@@

```{r t2m}
t2m <- ERA5_T2M() %>% 
  ReadNetCDF() %>% 
   normalise_coords() %>%
  .[, w := season_weights(time)] %>% 
  .[, .(t2m = mean(t2m*w)), by = .(lon, lat, time = seasonally(time))]
```

```{r rotated-regressions} 
rotate_reim <- function(real, imaginary, angle = 0) {
  z <- complex(real = real, imaginary = imaginary)
  z <- rotate(z, angle)
  list(real = Re(z),
       imaginary = Im(z))
}

angles_regr <- -c(0, 45, 90, 45+90)*pi/180
angles_labs <- setNames(c("Real", "45º", "Imaginary", "135º"), angles_regr*180/pi)

hgt850 <- ERA5_geopotential() %>% 
  ReadNetCDF(vars = c(hgt = "z"), 
             subset = list(level = 850)) %>% 
   normalise_coords() %>%
  .[season(time) == "SON"] %>% 
  .[, w := season_weights(time)] %>% 
  .[, .(hgt = mean(hgt*w)), by = .(lon, lat, time = seasonally(time))]

hgt850 <- ceof[, eof[[1]]$left, by = .(season)] %>%
  .[season == "SON"] %>%
  sep_ReIm(format = "wide") %>%
  .[, hgt := NULL] %>%
  .[, `:=`(Real = as.numeric(scale(Real)), Imaginary = as.numeric(scale(Imaginary))),
    by = .(season, cEOF)] %>%
  .[hgt850, on = "time", allow.cartesian = TRUE]

hgt850_regr <- lapply(angles_regr, function(a) {
  hgt850 %>% 
    copy() %>% 
    .[, c("Real", "Imaginary") := rotate_reim(Real, Imaginary, a)] %>% 
    .[, `:=`(Real = Real/sd(Real), Imaginary = Imaginary/sd(Imaginary)), 
      by = .(season, cEOF)] %>%
    .[, FitLm(hgt, Imaginary, Real, se = TRUE), by = .(lon, lat, cEOF, season)] %>%
    rm_intercept() %>%
    .[term == "Real"] %>% 
    .[, term := factor_ReIm(term)] %>% 
    .[, angle := a*180/pi]
}) %>% 
  rbindlist()

sst_for_regression <- ceof[, eof[[1]]$left, by = .(season)] %>%
  # .[cEOF == "cEOF2"] %>% 
  copy() %>% 
  sep_ReIm(hgt, "wide") %>%
  .[, hgt := NULL] %>%
  .[sst, on = "time", allow.cartesian = TRUE] %>%
  .[season(time) == "SON"] %>% 
  na.omit()

sst_regr <- lapply(angles_regr, function(a) {
  sst_for_regression %>% 
    copy() %>% 
    .[, c("Real", "Imaginary") := rotate_reim(Real, Imaginary, a)] %>% 
    .[, `:=`(Real = Real/sd(Real), Imaginary = Imaginary/sd(Imaginary)), 
      by = .(season, cEOF)] %>%
    .[, FitLm(t, Imaginary, Real, se = TRUE), by = .(lon, lat, cEOF, season)] %>%
    rm_intercept() %>%
    .[term == "Real"] %>% 
    .[, term := factor_ReIm(term)] %>% 
    .[, angle := a*180/pi]
}) %>% 
  rbindlist()

psi_for_regression <-  ceof[, eof[[1]]$left, by = .(season)] %>%
  # .[cEOF == "cEOF2"] %>% 
  copy() %>% 
  sep_ReIm(hgt, "wide") %>%
  .[, hgt := NULL] %>%
  .[psi, on = "time", allow.cartesian = TRUE] %>%
  .[season(time) == "SON"] %>%
  .[lev == 200]

psi_regr <- lapply(angles_regr, function(a) {
  psi_for_regression %>% 
    copy() %>% 
    .[, c("Real", "Imaginary") := rotate_reim(Real, Imaginary, a)] %>% 
    .[, `:=`(Real = Real/sd(Real), Imaginary = Imaginary/sd(Imaginary)), 
      by = .(season, cEOF)] %>%
    .[, FitLm(psi.z, Real, Imaginary, se = TRUE),by = .(lon, lat, lev, cEOF, season)] %>%
    rm_intercept() %>%
    .[term == "Real"] %>% 
    .[, psi.z := Anomaly(estimate), by = .(cEOF, season, lev, lat)] %>%
    .[, c("fx", "fy") := shceof::WaveFlux(.SD, p = 200), 
      by = .(lev, cEOF, season, term)] %>%
    .[, term := factor_ReIm(term)] %>% 
    .[, angle := a*180/pi]
}) %>% 
  rbindlist()

pp_for_regression <- ceof[, eof[[1]]$left, by = season] %>%
  # .[cEOF == "cEOF2"] %>% 
  copy() %>% 
  sep_ReIm(hgt, "wide") %>%
  .[, hgt := NULL] %>%
  .[cmap, on = "time", allow.cartesian = TRUE] %>%
  .[season(time) == "SON"] %>% 
  na.omit() %>% 
  .[, pp := pp/sd(pp), by = .(lon, lat)]

t2m_for_regression <-  ceof[, eof[[1]]$left, by = season] %>%
  # .[cEOF == "cEOF2"] %>% 
  copy() %>% 
  sep_ReIm(hgt, "wide") %>%
  .[, hgt := NULL] %>%
  .[t2m, on = "time", allow.cartesian = TRUE] %>%
  .[season(time) == "SON"] %>% 
  na.omit()

t2m_regr <- lapply(angles_regr, function(a) {
  t2m_for_regression %>% 
    copy() %>% 
    .[, c("Real", "Imaginary") := rotate_reim(Real, Imaginary, a)] %>% 
    .[, `:=`(Real = Real/sd(Real), Imaginary = Imaginary/sd(Imaginary)), 
      by = .(season, cEOF)] %>%
    .[, FitLm(t2m, Imaginary, Real, se = TRUE),
      by = .(lon, lat, cEOF, season)] %>%
    .[term == "Real"] %>% 
    rm_intercept() %>% 
    .[, term := factor_ReIm(term)] %>% 
    .[, angle := a*180/pi]

}) %>%
  rbindlist()
```

```{r enso_cor}
enso_cor <-  ceof[, eof[[1]]$left, by = .(season)] %>%
  sep_ReIm() %>%
  .[, hgt := scale(hgt), by = .(cEOF, part)] %>%
  enso[., on = "time"] %>%
  na.omit() %>%
  .[, correlate(hgt, oni), by = .(cEOF, part)]
```

```{r spoke}
spoke <- function(data, coords) {
  radius <-  grid::unit(.15, "snpc")

```{r plot_sst_psi}
plot_sst_psi <- function(sst_regr, psi_regr) {
  sst_gdata <- sst_regr %>%
    # .[cEOF == "cEOF2"] %>% 
    .[lat %between% c(-90, 10)] %>%
    .[, p_val := Pvaluate(estimate, std.error, df, "fdr"), 
      by = .(cEOF, term, angle)] %>%
    .[, var := "SST"]

  rsst <- max(range(abs(sst_regr[!is.na(cEOF), ]$estimate), na.rm = TRUE))
  breaks_sst <- AnchorBreaks(0, bins = 20, exclude = 0)(c(-rsst, rsst))

  stream_gdata <- psi_regr %>%
    # .[cEOF == "cEOF2"] %>% 
    # .[abs(estimate) > 1, estimate := NA] %>%
    .[lat %between% c(-90, 10)] %>%
    .[, p_val := Pvaluate(estimate, std.error, df, "fdr"), by = .(cEOF, angle, term)] %>%
    .[, var := "Streamfunction"]

  rstream <- max(range(abs(psi_regr$estimate), 
                       na.rm = TRUE))*1e-7
  breaks_stream <- AnchorBreaks(0, bins = 20, exclude = 0)(c(-rstream, rstream))

  # browser()
  stream_gdata %>% 
    periodic(lon = c(0, 360)) %>%
    ggplot(aes(lon, lat)) +

      geom_contour_fill(data = sst_gdata,
                      aes(z = estimate, fill = ..level..),
                      breaks = breaks_sst) +
    geom_contour_pval(data = sst_gdata, aes(z = p_val)) +

    scale_fill_divergent_discretised("SST (K)",
                                     labels = one_less,
                                     guide = guide_colorsteps_bottom(width = 15,
                                                                     show.limits = FALSE, ticks = TRUE,
                                                                     order = 1)) +
    ggnewscale::new_scale_fill() +
    geom_contour_fill(aes(z = estimate*1e-7, fill = ..level..),
                      breaks = breaks_stream) +
    geom_contour_pval(aes(z = p_val)) +
    # geom_streamline(aes(dx = fx, dy = fy), min.L = 3, L = 10,
    #                 size = 0.1, res = 2, nx = 22, ny = 12, 
    #                 arrow.angle = 14, arrow.length = 0.2) +
    geom_vector(aes(dx = fx, dy = fy), skip = 3, min.mag = 0.002, 
                size = 0.1) +
    scale_mag(max_size = .5, guide = "none") +
    geom_qmap(keep = .015) +
    scale_fill_divergent_discretised(latex2exp::TeX("Streamfunction ($m^2s^{-1}\\times 1e^{-7}$)"), 
                                     labels = one_less,
                                     low = scales::muted("red"),
                                     high = scales::muted("blue"),
                                     guide = guide_colorsteps_bottom(15, show.limits = FALSE,
                                                                     order = 99)) +
    grid_panel(spoke, aes(angle = -angle, var = "SST"), inherit.aes = FALSE) +
    scale_x_continuous(NULL, expand = c(0, 0), 
                       breaks = seq(-180, 180-60, by = 60),
                       labels = LonLabel) +

    scale_y_continuous(NULL, breaks  = seq(-90, 0, by = 30), expand = c(0, 0)) +

    facet_grid(angle ~ var, labeller = labeller(angle = angles_labs)) +
    # coord_quickmap(ylim = c(NA, 10)) +
    coord_sf(ylim = c(-89, 10), expand = FALSE,
             crs = "+proj=longlat +pm=180 +over",
             default_crs = "+proj=longlat +lon_wrap=180 +over") + 
    theme(legend.box = "horizontal",
          legend.text = element_text(size = rel(0.7)), 
          legend.box.spacing = grid::unit(0, "lines") ) +
    tag_facets("rc") 
}
```

```{r  sst-psi-2,  fig.cap = "(ref:sst-psi-2-cap)", fig.width = 7, fig.height=5.6, dependson="plot_sst_psi", fig.env="figure*"}
plot_sst_psi(sst_regr[cEOF == "cEOF2"], psi_regr[cEOF == "cEOF2"]) 
```

(ref:sst-psi-2-cap) Regression maps of cEOF2 with SST (K, column 1) and streamfunction zonal anomalies ($m^2/s\times10^-7$, shaded) with their corresponding activity wave flux (vectors) (column 2) for the 1979 -- 2019 period and different phases of the cEOF (illustrated in the lower-left arrow). 
Areas marked with dots have p-values smaller than 0.01 adjusted for FDR.

The connections between cEOFs and Tropical sources are also assessed.
Figure \@ref(fig:sst-psi-2) shows the regression maps of Sea Surface Temperatures (SST) and streamfunction anomalies at 200 hPa respectively upon standardised cEOF2, which can be interpreted as the anomalies associated with a 1 standard deviation increase of each index.
Besides showing the regression of the real and imaginary parts, we also show regressions in two intermediate directions. 
The Imaginary cEOF2 is associated with strong positive SST anomalies on the Central Pacific and negative anomalies over an area across the North of Australia and New Zealand, the South Pacific Convergence Zone (SPCZ) (Figure \@ref(fig:sst-psi-2).b1).
This pattern is almost canonically positive ENSO [@bamston1997] and indeed, the correlation between the Imaginary cEOF2 and the Oceanic Niño Index [@bamston1997] is significant and very high `r enso_cor[cEOF == "cEOF2" & part == "Imaginary"]$text`. 
On top of this ENSO-like pattern, there are positive anomalies in the western Indian Ocean and negative values in the eastern Indian Ocean, resembling a positive Indian Ocean Dipole.

Fittingly, the Imaginary cEOF2 is also associated with strong wave-like streamfunction anomalies emanating from the tropics (Figure \@ref(fig:sst-psi-2).b2), both from the Central Pacific sector and the Incean Ocean.
These are consistent with the effect of ENSO and the Indian Ocean Dipole on the extratropics: the SST anomalies initiate anomalous tropical convection that excites Rossby waves that propagate meridionally towards higher latitudes [@mo2000; @cai2011; @nuncio2015].

Not all phases of the cEOF2 are associated with the same tropical SSTs, however. Figure \@ref(fig:sst-psi-2).d1 and d2 show that the Real cEOF2 is not associated either with any significant SST nor streamfunction anomalies in the tropics.
The correlation between the Real cEOF2 and ENSO is also not significant (`r enso_cor[cEOF == "cEOF2" & part == "Real"]$text`). 
Rows a and c in Fig. \@ref(fig:sst-psi-2) show that the intermediate phases are still associated with significant SST anomalies over the Pacific Ocean, but with with slightly different location.
The 135º phase is associated with SST in the central Pacific, while the 45º phase is associated with SST in the eastern Pacific, which correspond roughly to the Eastern Pacific and Central Pacific "flavours" of ENSO (Citas).

(ref:enso-phase-cap) ONI plotted against cEOF2 phase for th 1979 -- 2019 period.
Years with magnitude of cEOF2 greater or smaller than the 50th percentile are shown as orange diamonds and green circles and respectively. 
Black line is the fit ONI ~ sin(phase) computed by weighted OLS using the magnitude of the cEOF2 as weights.

```{r enso-phase, fig.cap = "(ref:enso-phase-cap)", fig.width = width_column, fig.height=3, include = TRUE}
gdata <- ceof[, eof[[1]]$left, by = .(season)] %>%
  .[cEOF == "cEOF2"] %>% 
  # .[, hgt := Anomaly(hgt)] %>% 
  .[, arg := Arg(hgt)] %>%
  .[, mag := Mod(hgt)] %>%
  .[enso, on = "time"] %>%
  na.omit() %>%
  .[, q := ifelse(mag >= quantile(mag, 0.5), "|cEOF2| > 50%", "|cEOF2| < 50%")]

@@

  geom_smooth(data = ~qwrap(.x, arg = c(-pi, pi) ~ c(-2*pi, 2*pi)),
              method = "lm", aes(weight = mag),
              formula = y ~ I(sin(x)) -1,
              se = FALSE, color = "black", size = 0.5) +
  geom_point(aes(color = q,
                 shape  = q),
             size = 3) +
  ggrepel::geom_text_repel(data = ~.x[order(-oni)][1:3],
                           aes(label = year(time))) +

@@

The correlation between the absolute magnitude of the ONI and the magnitude of the cEOF2 is `r cor_enso_mag$text`.
This relationship, however, appears to be driven only by the three years with strongest ENSO events in the period (`r knitr::combine_words(year(gdata[order(-oni)][1:3]$time))`) which also coincide with the three years with strongest cEOF2 magnitude.
If those years are removed, the correlation becomes non-significant (`r cor_enso_mag_outlier$text`).
Furthermore, even when using all years, the Spearman correlation --which is robust to outliers-- is also non-significant (`r signif(cor_enso_spearman$estimate, 2)`, p-value`r report_p(cor_enso_spearman$p.value)`).
Therefore, although the location of tropical SST seem to have an effect in defining the phase of the cEOF2, the relationship between the magnitude of the cEOF2 train wave and ONI remains uncertain.

It could be concluded that the wave train represented by cEOF2 can be both part of the internal variability of the extratropical atmosphere or forced by tropical SSTs.
In the former case, the wave train has little phase preference.
However, when cEOF2 is excited by tropical SST variability, it tends to remain locked to the imaginary phase.
This explains the relative over-abundance of years with cEOF2 near positive and negative imaginary phase in Figure \@ref(fig:phase-histogram).

Unlike the cEOF2 case, there is no significant pattern of SST anomalies associated with either the Real or Imaginary cEOF1 (Figure \@ref(fig:sst-psi-1)).
Consistently, streamfunction anomalies do not show any tropical influence.
Instead, the real and Imaginary cEOF1 are associated with zonally wave activity fluxes in the extra-tropics around 60ºS, except for an equatorward flow from the coast of Antarctica around 150ºE in the real component.
This suggests that its variability is driven primary by the internal variability of the extra-tropics.

## Precipitation and 2-metre temperature {#precipitation}

@@

```{r plot_pp}
plot_pp <- function(pp_regr, t2m_regr, hgt850_regr) {
  lats <- c(-55, 10)
  breaks_pp <- AnchorBreaks(exclude = 0)(c(-.9, .9))

  pp_regr <- pp_regr %>% 
    copy() %>% 
    .[, var := "Precipitation"] %>% 
        .[, p.value := Pvaluate(estimate, std.error, df, "none"), 
      by = .(cEOF, angle)]

  t2m_regr <- t2m_regr %>% 
    copy() %>% 
    .[, var := "2m Temperature"] %>% 
    .[, p.value := Pvaluate(estimate, std.error, df, "none"), by = .(cEOF, angle)]

  pp_regr %>%
    # .[lat %between% lats] %>%

    periodic(lon = c(0, 360)) %>%
    ggplot(aes(lon, lat)) +

    geom_contour_fill(aes(z = estimate, fill = ..level..),
                      breaks = breaks_pp) +
    geom_contour_tanaka(aes(z = estimate), breaks = breaks_pp) +

    scale_fill_divergent_discretised("Precipitation (correlation)", 
                                     # limits = c(-1, 1),
                                     low = "#8E521C",
                                     high = "#00665E",
                                     guide = guide_colorsteps_bottom(15, order = 99)) +
    geom_contour_pval(aes(z = p.value)) +

    ggnewscale::new_scale_fill() +

    geom_contour_fill(data = t2m_regr, 
                      aes(z = estimate, fill = ..level..), 
                      breaks = AnchorBreaks(exclude = 0)) +
    geom_contour_tanaka(data = t2m_regr, 
                        aes(z = estimate),
                        breaks = AnchorBreaks(exclude = 0)) +

    scale_fill_divergent_discretised("Temperature",
                                     # limits = c(-1, 1),
                                     guide = guide_colorsteps_bottom(15, order = 1)) +
    geom_contour_pval(data = t2m_regr, aes(z = p.value)) +

    geom_contour2(data = copy(hgt850_regr)[, var := "2m Temperature"], 
                  aes(z = estimate, linetype = factor(-sign(..level..))), 
                  size = 0.2, breaks = AnchorBreaks(0, exclude = 0)) +

    grid_panel(spoke, aes(angle = -angle, var = "SST"), inherit.aes = FALSE) +
